Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-10-20 paint-contour-tree.tex
updates to surgery theorem

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

We define a ``cut" operation on manifolds that cuts along a contour to create
a manifold with a new boundary. Intuitively, any cut creates two disjoint boundaries, one
above and one below.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\cut(\MM,\phi)$}

\smallskip
\begin{asparaenum}
%	\item Insert the vertices of $\phi$ in $\MM$.
	\item Take an arbitrary edge $e$ whose interior intersect $\phi$.
	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
	at each intersection point of these contours with edges of the manifold. Triangulate all new faces.
	\item Delete the vertices (and incident edges, faces, etc.) of $\phi$ from the manifold.
\end{asparaenum}
\end{minipage}}

\medskip
This results in a new manifold which is boundary critical if the input manifold was. There are two new disjoint
boundary faces added. The manifold may be disconnected. Abusing notation, we use $\cut(\MM,\Psi)$
for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Psi$.

We now describe a procedure that converts the Reeb Graph of $\cut(\MM,\phi)$ to the Reeb Graph of $\MM$.

@@

\smallskip
\begin{asparaenum}
	\item Let $\MM' = \cut(\MM,\phi)$.
	\item Construct $\reeb(\MM')$ and let $B, C$ be the vertices corresponding to the new boundaries created
	in $\MM'$. (One is a minima and the other is maxima.)
	\item Since $B, C$ are leaves, they each have unique neighbors $B'$ and $C'$, respectively. Insert
	edge $(B',C')$ and delete $B,C$ to obtain a new graph.
\end{asparaenum}
\end{minipage}}

\medskip
\begin{lemma} \label{lem:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
\end{lemma}

\begin{proof} The only difference between $\reeb(\MM)$ and $\reeb(\cut(\MM,\phi))$ is in the equivalence
class corresponding to $\phi$. This class is split into two classes in $\cut(\MM,\phi)$, which are joined
back by $\surgery(\MM,\phi)$.
\end{proof}

@@

In this section, we assume that $\MM$ is homeomorphic to $\SSS^2$ with boundaries.

\begin{definition} \label{def:dom} A manifold is \emph{extremum dominant} if there exists
a maximum (resp. minimum) $x$ such that every non-maximal (resp. non-minimal) \emph{vertex} in the manifold has an increasing (resp. decreasing) path to $x$.
\end{definition}

We will describe a linear time algorithm that partitions $\MM$ into a set of extremum dominant manifolds.
The first part of this requires a ``raining" procedure. Start at some point $x \in \MM$ and imagine rain at $x$.
The water will flow downwards along descending paths and ``wet" all the points encountered.

\begin{definition} \label{def:wet} The set of points $y$ such that there is a descending path from $x$ to $y$
is denoted by $\wet(x,\MM)$. A point $z$ is on the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
has non-trivial intersection with $\wet(x,\MM)$.
\end{definition}

Note that $\wet(x,\MM)$ (and its interface) can easily be computed in linear time in the size of the wet sub-manifold, using a descending BFS from $x$.

\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of disjoint, non-regular contours.
Furthermore, each such contour contains a merge vertex.
\end{claim}

\begin{proof}
 First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
 as one can follow the descending path to $y$ and then walk along the contour.

 So let $y$ be a point in the interface.
 For $\eps$ sufficiently small, the points in any $\eps$ neighborhood of $y$ that lie below $y$ must have a descending path from $y$ and hence must be wet.
 Therefore in order for $y$ to have non-trivial intersection with $\wet(x,\MM)$ for any arbitrarily small $\eps$ there must be a dry point which lies 
 strictly above $y$.  In particular, this implies there must be an entire dry contour which lies just above $y$.

@@

 By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
 and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
\end{proof}

@@

We will define a \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \eps$. Let $\lift(y) = \cont(z)$.

\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
results in two disjoint manifolds, one consisting entirely of dry points.
\end{claim}

\begin{proof} Note that $\lift(\phi)$ is a closed, dry curve. By the Jordan Curve Theorem,
$\cut(\MM,\lift(\phi))$ results in two disjoint manifolds. Let $\NN$ be the manifold containing
the point $x$ that we called $\wet(x,\MM)$ from, and let $\NN'$ be the other manifold. 
If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
contradicting the fact that $\phi$ is dry.
\end{proof}

We now define the main partitioning procedure that cuts up a manifold $\NN$ homeomorphic to $\SSS^2$ with boundaries into a series of extremum
dominant manifolds. It takes as input the manifold and a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\rain(x,\NN)$}

\smallskip
\begin{compactenum}
	\item Compute the interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$, and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Call $\cut(\phi'_i,\NN_1)$ to get the dry manifold $\LL_i$ and wet manifold $\NN_{i+1}$.
		\item Let the new boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$.
\end{compactenum}
\end{minipage}}

\medskip
For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.

\begin{lemma} \label{lem:rain-1} Each output manifold $\MM_i$ is extremum dominant.
\end{lemma}

\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ consists of wet points. 
Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
boundaries are maxima.
\end{proof}

The procedure $\rain$ creates new boundary vertices on the edges it cuts each time it call the 
subroutine $\cut$.  The key to understanding the running time of this procedure is to bound the 
number of newly created vertices, for which we have the following lemma.

\begin{lemma}\label{lem:new-verts}
The total number of boundary vertices created by a call to $\rain(x, \NN)$ is $O(n)$, where $n=|\NN|$
\end{lemma}
\begin{proof}
As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
The wet part is in $\MM$ and will never be involved in any recursive call.

The dry part of $e$ (call it $e'$) needs to be considered.
A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
\end{proof}

As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.

@@

Consider the set of manifolds output by $\rain(x,\MM)$, there is a natural tree ordering defined over these manifolds where the 
root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).

%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
%the contour appears as a boundary maxima/minima.

\begin{claim} \label{clm:rain-reeb} Let $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
Then $\reeb(\MM)$ can be computed in $O(n)$ time.
\end{claim}

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\section{Contour tree basics} \label{sec:basics}

We detail the basic definitions about contour trees. Most of this is taken from Chapter 6 of Carr's thesis~\cite{Carr}.
Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a simplicial complex in $\RR^d$. 
All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated
\emph{boundary} simplices. \Sesh{Some mention of representation and data structures here.}
It is standard to assume
that $f$ is Morse (we explain later what this entails).

Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$.
We will often omit $f$, and assume that $f$ is encoded in the representation of $\MM$.

\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x \in \MM| f(x) = h\}$.
A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
\end{definition}

Not that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$.

\begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours that intersect $B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours that intersect $B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.
\end{definition}

The definition allows us to categorize points in $\MM$ depending on the local changes in topology.

\begin{definition} \label{def:points} A point $x \in \MM$ is categorized as follows.
\begin{asparaitem}
	\item Regular: Both up-degree and down-degrees are $1$.
	\item Maximum: Up-degree is $0$.
	\item Minimum: Down-degree is $0$.
	\item Join: Up-degree is strictly greater than $1$.
	\item Split: Down-degree is strictly greater than $1$.
\end{asparaitem}
Non-regular points are called \emph{critical}. Joins and splits are called \emph{saddles}.
Maxima and minima are called \emph{extrema}.
\end{definition}

We will enforce a special constraint on boundaries, encapsulated in the following definition.

\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of $F$ not in $F$ either have function values strictly greater or strictly less
than this value.
\end{definition}

We use $\cV$ to denote the set of critical points in $\MM$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(x) = h$, for some $x \in \cV$. The Morse assumption
implies that all critical points have distinct values. A contour that involves a critical
point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
Critical values include these boundary values. By boundary criticality, all contours are themselves
represented as $d-1$-dimensional simplicial complices.

\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
and we can partition the regular contours into these classes. Every such class maps to intervals
of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
to be created at $x_i$ and destroyed at $x_j$.

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventially, edges are directed from higher function value to lower.)
\end{definition}

(It is not immediately obvious that this graph is a tree. But it is equivalent to the Reeb graph
of $f:\MM \rightarrow \RR$, which is known to be a tree.) We denote the contour tree of $\MM$
by $\reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.

We use a key theorem from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.

\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree to which $P$
maps, and for every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ that maps to $Q$.
\end{theorem}

\section{Divide and conquer through contour surgery} \label{sec:surgery}

{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
a new simplicial complex with an added boundary. Given contour $\phi$, this is constructing the simplicial complex $\MM \setminus \phi$. 
We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
the contour at value $f(\phi^+) + \eps$ (resp. $f(\phi^-) - \eps$), which is at distance $\eps$
from $\phi$.

An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$. (Think of the $d+1$-dimension
as height.) Given some point $x$
on an $h$-contour $\phi$ (with some simplex that $x$ is present in), we can walk along $\MM$ from $x$ and determine $\phi$.
We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
by splitting every facet $F$ that $\phi$ intersects into an ``upper" facet and ``lower" facet. Algorithmically,
we cut facet $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper facet. Analogously, we cut with $\phi^-$ to get the lower facet.
The facets are then triangulated to ensure that they are all simplices.
Given that $\phi$ cannot cut a boundary and all non-boundary facets
have constant size, we omit the algorithmic description for this process. We only note that this
creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
This new simplicial complex is denoted by $\cut(\phi,\MM)$.
This can be constructed in time linear in $|\phi|$.

%We denote this new manifold by $\cut(\phi,\MM)$, and the new boundaries by $\phi^+$ and $\phi^-$.
%Abusing notation, we use $\cut(\phi,\MM)$
%for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Phi$.

\ignore{
\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\cut(\phi,\MM)$}

\smallskip
\begin{asparaenum}
%	\item Insert the vertices of $\phi$ in $\MM$.
	\item For every edge $e$ that intersects $\phi$ at point $x$: add edges $(e,x^+)$ and $(e,x^-)$
	where $x^+$ (resp. $x^-$) is infinitesimally above (resp. below) $x$.
	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
	at each intersection point of these contours with edges of $\MM$. Add  
	\item Delete the edges between $\phi_x$ and $\phi_y$.
\end{asparaenum}
\end{minipage}}

\medskip
This results in a new boundary critical simplicial complex. There are two new disjoint
boundary simplices added. The manifold may be disconnected. 
}

We now describe a high-level approach to construct $\reeb(\MM)$.

@@

\smallskip
\begin{asparaenum}
	\item Let $\MM' = \cut(\phi,\MM)$.
	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
	in $\MM'$. (One is a minima and the other is maxima.)
	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
\end{asparaenum}
\end{minipage}}

\medskip
\begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
\end{theorem}

The main theorem is a direct consequence of the following lemma.

\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contain in contour class (edge of $\reeb(\MM))$
$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup (\cV(\MM) \setminus \phi)$
and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
\end{lemma}

\begin{proof} The contours of $\MM'$ are exactly the contours of $\MM$ with $\phi^+$ and $\phi^-$, without $\phi$.
All new vertices created in $\MM'$ are present on $\phi^+$ and $\phi^-$. That proves the first part.

Any contour class in $\MM'$ (edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM)$.

Consider the contour class corresponding to edge $(u,v)$ of $\cC(\MM)$. There is a natural ordering
of the contours by function value, ranging from $f(u)$ to $f(v)$. All contours in this class ``above" $\phi$
form a maximal contour class in $\MM'$, represented by edge $(u,\phi^+)$. Analogously, there is another
contour class represented by edge $(\phi^-,v)$. We have now accounted for all contours in $\cC(\MM')$,
completing the proof.
\end{proof}

A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
two disconnected complices.

\begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
disconnected simplicial complices.
\end{theorem}

\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
in $\MM' = \cut(\phi,\MM)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$. Contradiction.
To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
\end{proof}

@@

In this section, we describe a linear time procedure that partitions $\MM$ into special
\emph{extremum dominant} simplicial complices.

\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has decreasing path to $x$.
Analogously define \emph{maximum dominant}. 
\end{definition}

The first aspect of the partitioning is ``raining". Start at some point $x \in \MM$ and imagine rain at $x$.
The water will flow downwards along descending paths and ``wet" all the points encountered. Note that this procedure considers all points
of the manifold, not just vertices.

\begin{definition} \label{def:wet} Fix $\MM$ and $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
is denoted by $\wet(x,\MM)$ (which is in turn represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
has non-trivial intersection with $\wet(x,\MM)$.
\end{definition}

The following claim gives a description of the interface. While its meaning is quite intuitive, the proof is tedious.

\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of contours,
each containing a join vertex.
\end{claim}

\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
 (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.) The converse is also true,
 so $\wet(x,\MM)$ contains entire contours.

Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)$ is dry.)
The points in $B_\eps(y)$ that lie below $y$ have a descending path from $y$ and hence must be wet.
There must be a dry point in $B_\eps(y)$ that is above $y$, and hence,
there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.

Suppose $y$ is in the interface. Consider the set $\Gamma_y$ of points connected to $y$
by a path with all points at height $f(y)$. (This is a set of $f(y)$-contours.) Suppose for contradiction
that $\forall z \in \Gamma_y$, $z$ has up-degree $1$. Consider the non-ascending path from $x$ to $y$ and let $z$
be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
have up-degree $1$ (ref. to \Def{deg}), then $\psi$ is the unique $f(y)+\delta$-contour
intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\phi$ is wet
and $\psi$ is dry.

Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
lies in $\Gamma_y$ (which is wet). Hence, this gives the interface, which is a contour containing
the join $w$.
\end{proof}

\ignore{
\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
 (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.)
 First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
 as one can follow the non-ascending path to $y$ and then walk along the contour.

 So let $y$ be a point in the interface.

@@

 By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
 and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
\end{proof}
}

Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
We perform a non-ascending walk from $x$. Any facet $F$ of $\MM$ encountered has some (if not all) portion of in $\wet(x,\MM)$.
This portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.

@@

We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through regular point $z$. Note that $\lift(\phi)$ is dry.

\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
results in two disjoint simplicial complices, one consisting entirely of dry points.
\end{claim}

\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complices. Let $\NN$ be the complex containing
the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
%
%
%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
%contradicting the fact that $\phi$ is dry.
\end{proof}

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complices. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
of the procedure is that the rain alternately flows downwards and upwards. In other words, when
we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
opposite direction. While this is way of implementing $\rain$, it is conceptually easier
to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
is easily achieved by just negating the height values. (From a running time standpoint, it suffices
to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

%
%We critically perform
%an \emph{inversion} step, which is simply achieved by negating
%
\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\rain(x,\NN)$}

\smallskip
\begin{compactenum}
	\item Determine interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\phi'_i,\NN_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$ together with any complices output by recursive calls.
\end{compactenum}
\end{minipage}}

\medskip
For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
We first argue the correctness of $\rain$.

\begin{lemma} \label{lem:rain-1} Each output $\MM_i$ is extremum dominant.
\end{lemma}

\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
boundaries are maxima.
\end{proof}

Focus on $\rain(x,\MM)$. As $\rain$ proceeds, new facets/simplices are created because of repeated cutting. Indeed, any facet $F$
in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created facets, for which we have the following lemma.

\begin{lemma}\label{lem:new-verts}
A facet $F \in \MM$ is cut at most once during the $\rain(x,\MM)$.
\end{lemma}
\begin{proof} All notation here follows that in the pseudocode of $\rain$.
Firstly, by \Thm{jordan}, all the pieces of which $\rain$ is invoked are disjoint.
Secondly, all recursive calls are made on dry complices.

Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
Specially, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
Since the cut is infinitesimally above an interface, the lower portion is basically wet.
It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
All in all, $F$ is cut at most once.
\ignore{
As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
The wet part is in $\MM$ and will never be involved in any recursive call.

The dry part of $e$ (call it $e'$) needs to be considered.
A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
}
\end{proof}

\begin{theorem} \label{thm:rain-time} The total running time of $\rain(x,\MM)$ is $O(|\MM|)$.
\end{theorem}

\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
complexities of the wet complices and the new boundaries created. No facet is every wet twice,
and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
bounds the complexities of the boundaries created.
\end{proof}

\ignore{
As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.

@@

%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
%the contour appears as a boundary maxima/minima. 
}

\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
%be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
$\reeb(\MM)$ can be constructed in $O(|\MM|)$ time.
\end{claim}

\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
labeled with some $\MM_i$.
%
%There is a natural tree ordering defined over these manifolds where the 
%root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
%(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
%$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).  
Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
contour tree of its children in the tree using the $\surgery$ procedure. 
Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
%
%
%This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).  The correctness follows by induction and  \Lem{surgery}.
\end{proof}

\section{Painting to compute contour trees} \label{sec:paint}
